Kinematics of the Palomar 5 stellar stream from RR Lyrae stars

### 2019-09-05 paper/main.tex
Updates from Overleaf

Using a small number of RRLs then thought to be associated with Pal 5, \citet{Vivas:2006} measured a distance to the cluster of $d = 22.3~\kpc$.
However, this used a now imprecise $M_V$--$\feh$ relation to compute the distance modulus.
Adopting $\feh = -1.35$ \citep{Ishigaki:2016} and using the updated RRL $M_V$--$\feh$ relation from \citet{Muraveva:2018}, we find instead that this past RRL-based distance to Pal 5 is instead $d = 21.2 \pm 0.6~\kpc$.

@@

% \todo{APW: still typical?}
%CM: yes, Nab/Nc has a long tail towards large values
Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find \NRRcl\ RRLs in the cluster itself (2 type~\typeab\ and 8 type~\typec) and \NRRtails\ in the cluster's tails (13 type~\typeab\ and 4 type~\typec), all with membership probabilities larger than 0.5. In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.25$,a typical ratio found Oosterhoff I populations \citep{Smith:1995}.
\figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5.
\ab{can we provide a sentence for what it (physically) means to be an oosterhoff I population?}
The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram. The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.54~d$.
Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.4$ \citep{Dotter:2011}. Interestingly, there seems to be a strong segregation of the RRL types: all but two of the \rrab~are found in the cluster tails and two thirds of the \rrc~are found in the cluster itself.
\ab{this is interesting indeed -- can we provide a hypothesis?}
Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.

%. and the Sagittarius dwarf galaxy, but not in any other dwarf spheroidal galaxy. \todo{CM:( any other point to be made about this?)}. \sa{as someone who doesn't know much about RRLs, how weird is the above? maybe some more discussion/elaboration? } \cm{well, it's not weird at all. I've added a sentence saying so.}

@@

%\todo{APW: update numbers with new catalog}
To estimate the expected absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.  The $\log{N_{RR}}-M_V$ relation has a large scatter so, particularly for such low numbers of RRLs, it will only allow us to give rough limits on the cluster's absolute magnitude and, therefore, of its luminosity. Correcting the observed number of 27~RRLs by the median completeness of 92\%, we get a total expected number of 29~RRLs, for which we find a total luminosity $L_V=4.6_{-3.0}^{+8.8}\times 10^4 \mathrm{L}_\sun$.
We then estimate a stellar mass of $0.9_{-0.6}^{+1.8}\times 10^5 \msun$, by assuming a V-band mass-to-light ratio $M/L=2$ corresponding to a simple stellar population with the age (12 Gyr) and metallicity ($\FeH=-1.4$) of the cluster \citep{Dotter:2011}, computed from the \citet{Bruzual:2003} models \footnote{We have used Z=0.0004 ($\FeH=-1.6$), the nearest to the cluster's metallicity available in the models.}. Our mass estimate is larger than that of \citet{Ibata:2017}, who report a current mass of $1.2\pm 0.4 \times10^4\msun$ for the cluster and tails; although their estimate for the initial mass ($4.7\pm 1.5\times10^4\msun$) of the system is at the lower end of our confidence limits.

@@

\acknowledgments
This work was performed in part during the Gaia19 workshop and the 2019 Santa Barbara Gaia Sprint (also supported by the Heising-Simons Foundation), both hosted by the Kavli Institute for Theoretical Physics at the University of California, Santa Barbara. The Flatiron Institute is supported by the Simons Foundation. CM acknowledges funding from the MIA program at Universidad de la Republica, Uruguay, and is grateful for the hospitality and support of the Flatiron Institute where part of this research was carried out.

@@

\software{
    astropy \citep{astropy, astropy:2018},
    dustmaps \citep{dustmaps},
    emcee \citep{emcee},
    gala \citep{gala},
    IPython \citep{ipython},
    matplotlib \citep{mpl},
    numpy \citep{numpy},
    scipy \citep{scipy}
}
\clearpage
\appendix

>>>>

Using a small number of RRLs then thought to be associated with Pal 5, \citet{Vivas:2006} measured a distance to the cluster of $d = 22.3~\kpc$.
However, this used a now imprecise $M_V$--$\feh$ relation to compute the distance modulus.
Adopting $\feh = -1.35$ \citep{Ishigaki:2016} and using the updated RRL $M_V$--$\feh$ relation from \citet{Muraveva:2018}, we find that this past RRL-based distance to Pal 5 is instead $d_{\rm Pal 5} = 21.2 \pm 0.6~\kpc$.
To summarize: Our inferred distance to the Pal 5 cluster, $d_{\rm Pal 5} = \clderr$, is consistent with (but more precise than) all recent distance measurements, once corrected for dust extinction and updated calibrations of the horizontal branch absolute magnitude.

@@

% \todo{APW: still typical?}
%CM: yes, Nab/Nc has a long tail towards large values
Assuming a Jacobi radius of $11'$---computed assuming $\textrm{M}_{\rm Pal 5} \approx 1.2\times10^4~\msun$ \citep{Kuepper:2015} and using the Milky Way mass model in \citet{gala}---we find \NRRcl\ RRLs in the cluster itself (2 type~\typeab\ and 8 type~\typec) and \NRRtails\ in the cluster's tails (13 type~\typeab\ and 4 type~\typec), all with membership probabilities larger than 0.5. In total, we find the number ratio of RRL types to be $N_{ab} / N_{c} \approx 1.25$, a typical ratio found in Oosterhoff I populations \citep{Smith:1995},  comprised mainly of stars still in the zero-age horizontal branch. \figurename~\ref{fig:PA_diagram} shows the period-amplitude diagram for the RRL stars in Pal~5. The full sample of PS1 RRLs is shown in the background for comparison, as a grayscale density histogram. The distribution of the \typeab~RRLs clearly indicate that Pal~5 is an Oosterhoff I cluster, consistent with having a mean period of $0.54~d$. Our findings of the updated ratio of \rrab~to~\rrc, mean period, and Period-Amplitude distribution of the \rrab\ RRLs all support an Oosterhoff I classification for Pal~5, consistent with the cluster's metallicity, $\FeH=-1.4$ \citep{Dotter:2011}. 
Out of our newly found \rrab\ RRLs, three are High Amplitude Short Period (HASP, $P <0.48$~d, $A_V>0.75$~mag $=A_g>$) stars, again consistent with Pal~5's metallicity,  as this type of RRL is only found in relatively metal-rich globular clusters  \citep[$\FeH>-1.5$;][]{Monelli:2017}.

Interestingly, there seems to be a strong segregation of the RRL types: all but two of the \NRRab\ \rrab~are found in the cluster tails and two thirds of the \NRRc\ \rrc~are found in the cluster itself. At fixed metallicity and helium abundance, position along the horizontal branch is determined by (present-day) mass, with less massive stars ending up at lower temperatures as type \typeab~stars. 
The dichotomy of RRL types between the stream and cluster is therefore suggestive of being a signature of mass segregation: The (slightly) higher mass \rrc\ stars seem to be preferentially more bound to the cluster.
However, the mass difference between \rrab\ and \rrc\ stars is expected to be small, $\lesssim 0.1~\msun$, and so unlikely to have a significantly different dynamical history---this remains an intriguing puzzle.

@@

To estimate the expected total absolute magnitude of Pal 5 we follow the same procedure as in \citet{Mateu:2018}, who use a linear fit of the relation between the absolute magnitude $M_V$ and the number of \typeab\ RRL, based on data from Galactic dwarf galaxies and globular clusters, here including the RR\typec~stars.  The $\log{N_{RR}}-M_V$ relation has a large scatter so, particularly for such low numbers of RRLs, it will only allow us to give rough limits on the cluster's absolute magnitude and, therefore, of its luminosity. Correcting the observed number of 27~RRLs by the median completeness of 92\%, we get a total expected number of 29~RRLs, for which we find a total luminosity $L_V=4.6_{-3.0}^{+8.8}\times 10^4 \mathrm{L}_\sun$.
We then estimate a stellar mass of $0.9_{-0.6}^{+1.8}\times 10^5 \msun$, by assuming a V-band mass-to-light ratio $M/L=2$ corresponding to a simple stellar population with the age (12 Gyr) and metallicity ($\FeH=-1.4$) of the cluster \citep{Dotter:2011}, computed from the \citet{Bruzual:2003} models \footnote{We have used Z=0.0004 ($\FeH=-1.6$), the nearest to the cluster's metallicity available in the models.}. Our mass estimate is larger than that of \citet{Ibata:2017}, who report a current mass of $1.2\pm 0.4 \times10^4\msun$ for the cluster and tails; although their estimate for the initial mass ($4.7\pm 1.5\times10^4\msun$) of the system is at the lower end of our confidence limits.

@@

\acknowledgments

This work was performed in part during the Gaia19 workshop and the 2019 Santa Barbara Gaia Sprint (also supported by the Heising-Simons Foundation), both hosted by the Kavli Institute for Theoretical Physics at the University of California, Santa Barbara. The Flatiron Institute is supported by the Simons Foundation. CM acknowledges funding from the MIA program at Universidad de la Republica, Uruguay, and is grateful for the hospitality and support of the Flatiron Institute where part of this research was carried out.

@@

\software{
    astropy \citep{astropy, astropy:2018},
    dustmaps \citep{dustmaps},
    emcee \citep{emcee},
    gala \citep{gala},
    IPython \citep{ipython},
    matplotlib \citep{mpl},
    numpy \citep{numpy},
    scipy \citep{scipy}
}

\appendix
